---
title: "Introduction"
author: "Sebastian Alejos"
date: "2025-08-23"
output: html_document
---

## 👋 Hello Professor

My name is **Sebastian Alejos **
I am a **Queens Native**, with a passion for horror tropes and storytelling.
I am doing my masters in **Business Analytics with a concentration in data analytics**.

Currently, I am fulltime at a small company in Brooklyn that supplies routers for Verizon Representatives nationwide.

I want to make sure this files transfers correctly to my **github**, thus I will send another file throughout the semester with my linkedin and resume.



```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`



```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```